Searching for Software in NSF Awards

### 2023-02-10 paper/paper.qmd
Last 9 page paper push
Use ACM template for paper

---
title: "Searching for Software in NSF Awards"
author:
  - name: "Eva Maxfield Brown"
    orcid: 0000-0003-2564-0373
    affliations:
      - name: University of Washington Information School
  - name: "Lindsey Schwartz"
    orcid: XXXX-XXXX-XXXX-XXXX
    affliations:
      - name: University of Washington Information School
  - name: "Richard Lewei Huang"
    orcid: XXXX-XXXX-XXXX-XXXX
    affliations:
      - name: University of Washington Information School
  - name: "Nicholas Weber"
    orcid: XXXX-XXXX-XXXX-XXXX
    email: nmweber@uw.edu
    affliations:
      - name: University of Washington Information School

abstract: |
  Software production, use, and reuse is a crucial part of scholarly work but software products of research are in many cases not easily identifiable or discoverable. The first step of automatic linkage of research and software is identification. In this paper we present two datasets to study the identification and production of research software. The first dataset contains almost 1000 annotations of software production from National Science Foundation (NSF) awarded research projects. We use this dataset to train models which predict software production from either the text of an NSF award's abstract or project outcomes report. Our second dataset is created by applying the trained predictive models across the abstracts and project outcomes reports for all NSF funded projects between the years of 2010 and 2023. The result is an inferred dataset of software production for over XXXXX NSF awards. We release the NSF-Soft-Search dataset to aid in identifying and understanding research software production: https://github.com/si2-urssi/eager

@@

execute:
  echo: false
---

@@

from_abstract = model_results[
  model_results["predictive_source"] == "abstract-text"
].reset_index(drop=True)

Markdown(from_abstract.to_markdown(index=False))
```

@@

from_po = model_results[
  model_results["predictive_source"] == "project-outcomes"
].reset_index(drop=True)

Markdown(from_po.to_markdown(index=False))
```

@@

    desc_table_rows.append({
        "Funding Program": program,
        "# Awards Funded": total_awards_funded,
        "# 'Produce Software'": num_awards_will_produce_software,
        "% 'Produce Software'": percent_awards_will_produce,
    })

desc_table = pd.DataFrame(desc_table_rows).sort_values(
    by=["# Awards Funded", "% 'Produce Software'"],
    ascending=False,
).reset_index(drop=True)
Markdown(desc_table.to_markdown(index=False))
```

@@

        soft_produced_over_time_rows.append({
            "Funding Program": program_name,
            "Award Start Year": start_year,
            "% 'Produce Software'": predicted_count / award_count,
        })

@@

# Ensure that hue colors are consistent throughout plots
soft_produced_over_time = soft_produced_over_time.sort_values(
  by="Funding Program",
)

# Plot over time
ax = sns.barplot(
    data=soft_produced_over_time,
    x="Award Start Year",
    y="% 'Produce Software'",
    hue="Funding Program",
    order=[
        str(i)
        for i in range(
            high_soft_prog_df.startYear.astype(int).min(),
            high_soft_prog_df.startYear.astype(int).max(),
        )
    ],
)
sns.move_legend(
    ax,
    "lower center",
    bbox_to_anchor=(.5, 1),
    ncol=len(soft_produced_over_time["Funding Program"].unique()),
    title="Funding Program",
    frameon=False,
)
plt.xticks(rotation=45, ha="right")
plt.tight_layout()
```

@@

        soft_produced_over_duration_rows.append({
            "Funding Program": program_name,
            "Award Duration (Years)": award_duration,
            "% 'Produce Software'": predicted_count / award_count,
        })

@@

# Ensure that hue colors are consistent throughout plots
soft_produced_over_duration = soft_produced_over_duration.sort_values(
  by="Funding Program",
)

# Plot over time
ax = sns.barplot(
    data=soft_produced_over_duration,
    x="Award Duration (Years)",
    y="% 'Produce Software'",
    hue="Funding Program",
)
sns.move_legend(
    ax,
    "lower center",
    bbox_to_anchor=(.5, 1),
    ncol=len(soft_produced_over_time["Funding Program"].unique()),
    title="Funding Program",
    frameon=False,
)
plt.tight_layout()
```

* Number of examples that initially were predicted as producing software but didn't over time
* Number of examples that initially were not predicted to produce software but did over time

@@

::: {.content-visible unless-format="html"}
# References
:::

>>>>

---
title: "Searching for Software in NSF Awards"
author:
  - name: "Eva Maxfield Brown"
    email: evamxb@uw.edu
    orcid: 0000-0003-2564-0373
    affliation:
        name: University of Washington Information School
        city: Seattle
        state: Washington
        country: USA
  - name: "Lindsey Schwartz"
    email: lsschwar@uw.edu
    affliation:
        name: University of Washington Information School
        city: Seattle
        state: Washington
        country: USA
  - name: "Richard Lewei Huang"
    email: lwhuang@uw.edu
    affliation:
        name: University of Washington Information School
        city: Seattle
        state: Washington
        country: USA
  - name: "Nicholas Weber"
    email: nmweber@uw.edu
    orcid: 0000-0002-6008-3763
    affliation:
        name: University of Washington Information School
        city: Seattle
        state: Washington
        country: USA

abstract: |
  Software production, use, and reuse is a crucial part of scholarly work but software products of research are in many cases not easily identifiable or discoverable. The first step of automatic linkage of research and software is identification. In this paper we present two datasets to study the identification and production of research software. The first dataset contains almost 1000 annotations of software production from National Science Foundation (NSF) awarded research projects. We use this dataset to train models which predict software production from either the text of an NSF award's abstract or project outcomes report. Our second dataset is created by applying the trained predictive models across the abstracts and project outcomes reports for all NSF funded projects between the years of 2010 and 2023. The result is an inferred dataset of software production for over 150,000 NSF awards. We release the NSF-Soft-Search dataset to aid in identifying and understanding research software production: https://github.com/si2-urssi/eager

@@

execute:
  echo: false

keep-tex: true
keep-ipynb: true

# acm-specific metadata
acm-metadata:
  # comment this out to make submission anonymous
  anonymous: true

  # comment this out to build a draft version
  # final: true

  # comment this out to specify detailed document options
  acmart-options: sigconf, review

  # acm preamble information
  copyright-year: 2023
  acm-year: 2023
  copyright: acmcopyright
  doi: XXXXXXX.XXXXXXX
  conference-acronym: JCDL
  conference-name: Joint Conference on Digital Libraries
  conference-date: June 26--30, 2023
  conference-location: Santa Fe, NM
  # price: "15.00"
  # isbn: 978-1-4503-XXXX-X/18/06

  # if present, replaces the list of authors in the page header.
  shortauthors: Brown et al.

  keywords:
    - datasets
    - text classification
    - research software
---

@@

from_abstract = model_results[
  model_results["predictive_source"] == "abstract-text"
].reset_index(drop=True)
from_abstract.loc[:, from_abstract.columns != "predictive_source"]
```

@@

from_po = model_results[
  model_results["predictive_source"] == "project-outcomes"
].reset_index(drop=True)
from_po.loc[:, from_po.columns != "predictive_source"]
```

@@

    desc_table_rows.append({
        "Program": program,
        "# Awards": total_awards_funded,
        "# Software": num_awards_will_produce_software,
        "% Software": percent_awards_will_produce,
    })

desc_table = pd.DataFrame(desc_table_rows).sort_values(
    by=["# Awards", "% Software"],
    ascending=False,
).reset_index(drop=True)
desc_table
```

@@

        soft_produced_over_time_rows.append({
            "Program": program_name,
            "Award Start Year": start_year,
            "% 'Produce Software'": predicted_count / award_count,
        })

@@

# Ensure that hue colors are consistent throughout plots
soft_produced_over_time = soft_produced_over_time.sort_values(
  by="Program",
)

# Plot over time
ax = sns.barplot(
    data=soft_produced_over_time,
    x="Award Start Year",
    y="% 'Produce Software'",
    hue="Program",
    order=[
        str(i)
        for i in range(2010, 2023)
    ],
)
sns.move_legend(
    ax,
    "lower center",
    bbox_to_anchor=(.5, 1),
    ncol=len(soft_produced_over_time["Program"].unique()),
    title="Program",
    frameon=False,
)
plt.xticks(rotation=45, ha="right")
plt.tight_layout()
```

Using the NSF-Soft-Search Inferred dataset we can observe trends in software production over time. @fig-soft-over-time plots the percent of awards which we predict to have produced software (using the award's abstract) over time. While there are minor deviations year-to-year in predicted software production, we observe the "Math and Physical Sciences" (MPS) funding program as funding the most awards which we predict to produce software, with "Computer and Information Science and Engineering" (CISE), and "Engineering" (ENG) close behind.

@@

        soft_produced_over_duration_rows.append({
            "Program": program_name,
            "Award Duration (Years)": award_duration,
            "% 'Produce Software'": predicted_count / award_count,
        })

@@

# Ensure that hue colors are consistent throughout plots
soft_produced_over_duration = soft_produced_over_duration.sort_values(
  by="Program",
)

# We don't have many awards that are funded for more than 9 years
# This leads to deceptive plots -- remove greater than 9
soft_produced_over_duration = soft_produced_over_duration[
  soft_produced_over_duration["Award Duration (Years)"] <= 9
]

# Filter out oddities
soft_produced_over_duration = soft_produced_over_duration[
  soft_produced_over_duration["Award Duration (Years)"] >= 0
]

# Plot over time
ax = sns.barplot(
    data=soft_produced_over_duration,
    x="Award Duration (Years)",
    y="% 'Produce Software'",
    hue="Program",
)
sns.move_legend(
    ax,
    "lower center",
    bbox_to_anchor=(.5, 1),
    ncol=len(soft_produced_over_time["Program"].unique()),
    title="Program",
    frameon=False,
)
plt.tight_layout()
```

We can additionally observe trends in software production as award duration increases. @fig-soft-over-duration plots the percent of awards which we predict to have produced software (using the award's abstract) grouped by the award duration in years. We note that as award duration increases, the percentage of awards which are predicted to have produced software also tends to increase.

@@

# References
